### 2016-03-03 paper/icfp16-submission/semantics.tex
add base case
simplify 3.4 to just take a target expression
instead of top-level binder nonsense

\paragraph{Narrowing Types} The procedure % $\force{v}{t}{\vsu}{\tsu}$,
$$
\forcesym :: (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}
$$
defined in Figure~\ref{fig:narrow}, takes as input a value $v$, a type
$t$, and the current value and type substitutions, and refines $v$ to
have type $t$ by yielding a triple of either the same value and
substitutions, or yields the stuck state if no such refinement is
possible. In the case where $v$ is a hole, it first checks in the given
$\vsu$ to see if the hole has already been instantiated and, if so,
returns the existing instantiation.
\ES{explain the subtlety of narrowing a value hole to an uninstantiated type}
%
\begin{figure*}[ht]
$$
\begin{array}{lcl}
%% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
\forcesym                  & ::     & (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
% \force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
% \begin{cases}
%   \triple{v}{\vsu}{\tsu'} & \mbox{if } \tsu' = \unify{\{\typeof{v}, \thole\}}{\tsu} \\
%   \triple{\stuck}{\vsu}{\tsu} & \mbox{otherwise} \\

@@

\paragraph{Generating Values} The (non-deterministic) procedure $\gen{t}$
takes as input a type $t$ and returns a value of that type. For base
types, the procedure returns an arbitrary value. For functions, it returns
a lambda with a \emph{new} hole denoting the return value, and unconstrained
types (denoted by $\thole$) yield fresh holes constrained to have type \thole
(denoted by $\vhole{\thole}$). \ES{note that $\gen{\ttree{t}}{\tsu}$ will
always place holes in the generated nodes, they will be filled in later,
on demand.}

@@

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial input type upto trace $\trace$ will cause $f$ to get stuck
in \emph{at most} $k$ steps, where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
     \hastype{v}{t} and
     \tincompat{t}{\ptype{\trace}{f}},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
     in at most $k$ steps, where $k$ is the length of $\trace$.
\end{lem}
\begin{proof}
  By induction on $k$, the length of $\trace$. Let $s_{\trace} = \ptype{\trace}{f}$ and
  $\trace' = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e'}{\vsu'}{\tsu'}$
  such that $\step{e'}{\vsu'}{\tsu'}{e}{\vsu}{\tsu}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
  in at most $k$ steps.
%
  \begin{description}
  \item [Case \tincompat{t}{s_{\trace'}}:]
    The inductive hypothesis applies.
  \item [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
    Since $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$ we know
    that $s_{\trace'} \neq s_{\trace}$, and
    by Lemma~\ref{lem:force-inst} we know that we must have
    successfully called \forcesym at step $k$.
    %
    Lemma~\ref{lem:refine-partial} tells us
    $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
    specifically narrowed $s_{\trace'}$ to a type incompatible with $t$.
    %
    A case analysis of the evaluation rules shows that narrowing to
    $t$ instead cannot succeed.
  \end{description}
\end{proof}

@@

\paragraph{Inputs and Outputs}
%
The problem of generating inputs is undecidable in general.
%
Our witness generation procedure takes two inputs.
%
First, a search bound @n@ which is used to define the \emph{number} of
traces to explore. (We assume, without loss of generality, that all
traces are finite.)
%
Second, binder-expression tuples @[(f1,e1),...,(fn,en)]@
denoting the list of \emph{top-level definitions}:
%
\begin{code}
  let f1 = e1 in
  let f2 = e2 in
  ...
  let fn = en in
\end{code}
%
The witness generation procedure returns as output a list of (general)
witness expressions $e$ each of which is of the form @fn v1 ... vn@.
%
The \emph{empty} list is returned when no witness can be found after
exploring $n$ traces.
%
In what follows, we refer to @fn@ as the \emph{target} of witness generation.

\paragraph{Modeling Semantics}
%: \texttt{eval} and \texttt{isStuck}}
%
We resolve the non-determinism in the operational semantics
(\S~\ref{sec:semantics}) via the procedure:
%
\ES{HERE}
\begin{mcode}
  eval :: ($e$, $\vsu$, $\tsu$) -> [($v$, $\vsu$, $\tsu$)]
\end{mcode}
%
Due to the non-determinism introduced by \gensym, a call
\hbox{|eval($e$, $\vsu$, $\tsu$)|} returns the \emph{list} of possible results
of the form $\triple{e'}{\vsu'}{\tsu'}$ such that
$\steps{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$.
%
We determine whether a given value is stuck by using the predicate:
%
\begin{mcode}
  isStuck :: $v$ -> Bool
\end{mcode}

\paragraph{Generating Target Arguments}
%
We address the issue of currying by defining a procedure @genArgs@ such that \\
\hbox{@genArgs [(f1,e1),...,(fn,en)]@} yields a list of holes @[v1,...,vk]@
such that \hbox{@fn v1 ... vk@} \emph{does not} evaluate to a lambda.
%
This is achieved via a simple @loop@ that keeps adding holes to the
target application until @eval@uating the closed term yields a
non-lambda value.
%
The helper functions @witness@ and @close@ are used to respectively
apply the parameters to the target and close the result under the top-level
let-binders, prior to invoking \hbox{@eval@.}
%
That is, @mkApps@ creates a nested sequence of applications in
the usual left-associative style, and @mkLets@ takes a list of
binders and a body expression, and creates a sequence of nested
let-binders that close the body expression.

\begin{figure}[t]
\centering
\begin{mcode}
genArgs :: [($x$, $e$)] -> [$v$]
genArgs bs = loop []
  where
  loop vs  = case eval (close bs vs) of
               $\pair{\efun{x}{e}}{\_}$:_ -> loop ($\vhole{\thole}$ : vs)
               _        -> vs

close :: [($x$, $e$)] -> [$v$] -> ($e$, $\vsu$, $\tsu$)
close bs vs = (e, $\emptysu$, $\emptysu$)
  where
    e       = mkLets bs (witness bs vs)

witness :: [($x$, $e$)] -> [$v$] -> $e$
witness bs vs = mkApps target vs
  where
    target    = snd (last bs)
\end{mcode}
\caption{Find number of arguments for target function}
\label{fig:algo-gen-args}
\end{figure}

\paragraph{Generating Witnesses}
% Procedure \texttt{genWitness}}
%
Finally, Figure~\ref{fig:algo-gen-witness} summarizes the overall
implementation of our search for witnesses, in the procedure @genWitness@,
which takes as input a bound @n@ and a list of top level binders, and
returns a list of witness expressions $e$ that demonstrate how the input
program gets stuck.
%
The search proceeds as follows.
%
@(1)@ we invoke @genArgs@ to determine the number of parameters required
by the (curried) target function.
%
@(2)@ we obtain a closed expression $e$ by closing the target with the
computed holes.
%
@(3)@ we take the first @n@ traces returned by @eval@ on the closed $e$
and
@(4)@ we extract the substitutions corresponding to the $\stuck$ traces,
and use them to return the list of witnesses.
%
We obtain the following corollary of Theorem~\ref{thm:soundness}:

\begin{cor}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If @genWitness k [(f1,e1),...,(fn,en)]@ is non-empty,
  then for every type $t$ there exists $v$ of type $t$ such that
  $\steps{\eapp{f_n}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
\end{cor}

\begin{figure}[t]
\centering
\begin{mcode}
genWitness :: Nat -> [($x$, $e$)] -> [$v$]
genWitness n bs
       = [witness bs ($\vsu$ vs) | $\vsu$ <- $\vsu$s]
  where
   vs  = genArgs defs []        -- (1)
   cfg = close defs vs          -- (2)
   res = take n (eval cfg)      -- (3)
   $\vsu$s   = [$\vsu$ | $\triple{\stuck}{\vsu}{\tsu}$ <- res]  -- (4)
\end{mcode}
\caption{Generating Witnesses}
\label{fig:algo-gen-witness}
\end{figure}

>>>>

\paragraph{Narrowing Types} The procedure % $\force{v}{t}{\vsu}{\tsu}$,
$$
\forcesym :: (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}
$$
defined in Figure~\ref{fig:narrow}, takes as input a value $v$, a type
$t$, and the current value and type substitutions, and refines $v$ to
have type $t$ by yielding a triple of either the same value and
substitutions, or yields the stuck state if no such refinement is
possible. In the case where $v$ is a hole, it first checks in the given
$\vsu$ to see if the hole has already been instantiated and, if so,
returns the existing instantiation.
%
\begin{figure*}[ht]
$$
\begin{array}{lcl}
%% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
\forcesym                  & ::     & (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
% \force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
% \begin{cases}
%   \triple{v}{\vsu}{\tsu'} & \mbox{if } \tsu' = \unify{\{\typeof{v}, \thole\}}{\tsu} \\
%   \triple{\stuck}{\vsu}{\tsu} & \mbox{otherwise} \\

@@

\paragraph{Generating Values} The (non-deterministic) procedure $\gen{t}{\tsu}$
takes as input a type $t$ and returns a value of that type. For base
types, the procedure returns an arbitrary value. For functions, it returns
a lambda with a \emph{new} hole denoting the return value, and unconstrained
types (denoted by $\thole$) yield fresh holes constrained to have type \thole
(denoted by $\vhole{\thole}$). \ES{note that $\gen{\ttree{t}}{\tsu}$ will
always place holes in the generated nodes, they will be filled in later,
on demand.}

@@

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial input type upto trace $\trace$ will cause $f$ to get stuck
in \emph{at most} $k$ steps, where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
     \hastype{v}{t} and
     \tincompat{t}{\ptype{\trace}{f}},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
     in at most $k$ steps, where $k$ is the length of $\trace$.
\end{lem}
\begin{proof}
  By induction on $k$, the length of $\trace$. Let $s_{\trace} = \ptype{\trace}{f}$ and
  $\trace' = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e'}{\vsu'}{\tsu'}$
  such that $\step{e'}{\vsu'}{\tsu'}{e}{\vsu}{\tsu}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
  in at most $k$ steps.

  The base case where $k = 0$ (\ie empty traces) is trivial as no
  evaluation has happened and thus there can be no values that are
  incompatible with \ptype{0}{f}. So we proceed directly to the
  inductive case and split cases on whether $t$ is compatible with
  $s_{\trace'}$.
  %
  \begin{description}
  \item [Case \tincompat{t}{s_{\trace'}}:]
    The inductive hypothesis applies.
  \item [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
    Since $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$ we know
    that $s_{\trace'} \neq s_{\trace}$, and
    by Lemma~\ref{lem:force-inst} we know that we must have
    successfully called \forcesym at step $k$.
    %
    Lemma~\ref{lem:refine-partial} tells us
    $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
    specifically narrowed $s_{\trace'}$ to a type incompatible with $t$.
    %
    A case analysis of the evaluation rules shows that narrowing to
    $t$ instead cannot succeed.
  \end{description}
\end{proof}

@@

\paragraph{Inputs and Outputs}
%
The problem of generating inputs is undecidable in general.
%
Our witness generation procedure takes two inputs.
%
First, a search bound $k$ which is used to define the \emph{number} of
traces to explore. (We assume, without loss of generality, that all
traces are finite.)
%
% Second, binder-expression tuples @[(f1,e1),...,(fn,en)]@
% denoting the list of \emph{top-level definitions}:
% %
% \begin{code}
%   let f1 = e1 in
%   let f2 = e2 in
%   ...
%   let fn = en in
% \end{code}
% %
% The witness generation procedure returns as output a list of (general)
% witness expressions $e$ each of which is of the form @fn v1 ... vn@.
Second, the target expression $e$ that contains the type error, and may be a curried function of multiple arguments.
%
The witness generation procedure returns as output a list of (general)
witness expressions, each of which is of the form $e\ v_1 \ldots v_n$.
%
The \emph{empty} list is returned when no witness can be found after
exploring $k$ traces.
%
% In what follows, we refer to @fn@ as the \emph{target} of witness generation.

\paragraph{Modeling Semantics}
%: \texttt{eval} and \texttt{isStuck}}
%
We resolve the non-determinism in the operational semantics
(\S~\ref{sec:semantics}) via the procedure:
%
\begin{mcode}
  eval :: $\triple{e}{\vsu}{\tsu}$ -> [$\triple{v \cup \stuck}{\vsu}{\tsu}$]
\end{mcode}
%
Due to the non-determinism introduced by \gensym, a call
\hbox{|eval$\triple{e}{\vsu}{\tsu}$|} returns the \emph{list} of possible results
of the form $\triple{v \cup \stuck}{\vsu'}{\tsu'}$ such that
$\steps{e}{\vsu}{\tsu}{v \cup \stuck}{\vsu'}{\tsu'}$.
%
% We determine whether a given value is stuck by using the predicate:
% %
% \begin{mcode}
%   isStuck :: $v$ -> Bool
% \end{mcode}

\paragraph{Generating Target Arguments}
%
We address the issue of currying by defining a procedure @genArgs@ such that
\hbox{|genArgs $e$|} yields a list of holes |[$v_1$,...,$v_n$]|
such that $\eapp{e}{v_1 \ldots v_n}$ \emph{does not} evaluate to a lambda.
%
This is achieved via a simple @loop@ that keeps adding holes to the
target application until @eval@uating the term yields a
non-lambda value.
%
% The helper functions @witness@ and @close@ are used to respectively
% apply the parameters to the target and close the result under the top-level
% let-binders, prior to invoking \hbox{@eval@.}
%
% That is, @mkApps@ creates a nested sequence of applications in
% the usual left-associative style, and @mkLets@ takes a list of
% binders and a body expression, and creates a sequence of nested
% let-binders that close the body expression.

\begin{figure}[t]
% $$
% \begin{array}{lcl}
% \genArgs{e} &\defeq& \begin{cases} \genArgs{\addArg{e}},& \mbox{if } \triple{\efun{x}{e'}}{\vsu}{\tsu},\ldots = \doeval{e} \\
%  e,& \mbox{otherwise} \\
% \end{cases}
% \\
% \addArg{e} &\defeq& \eapp{e}{\vhole{\thole}}, \qquad \mbox{where } \ehole, \thole \mbox{ are fresh }
% \end{array}
% $$

\begin{mcode}
genArgs :: $e$ -> [$v$]
genArgs $e$ = loop []
  where
  loop vs = case eval $\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}$ of
             $\triple{\efun{x}{e}}{\vsu}{\tsu}$:_ -> loop ($\vhole{\thole}$ : vs)
                           -- $\ehole$, $\thole$ are fresh
             _          -> vs
\end{mcode}
\caption{Find number of arguments for target function}
\label{fig:algo-gen-args}
\end{figure}

\paragraph{Generating Witnesses}
% Procedure \texttt{genWitness}}
%
Finally, Figure~\ref{fig:algo-gen-witness} summarizes the overall
implementation of our search for witnesses, in the procedure @genWitness@,
which takes as input a bound @k@ and the target expression $e$, and
returns a list of witness expressions $\eapp{e}{v_1 \ldots v_n}$ that demonstrate how the input
program gets stuck.
%
The search proceeds as follows.
%
@(1)@ we invoke @genArgs@ to determine the number of parameters required
by the (curried) target expression.
%
% @(2)@ we obtain a closed expression $e$ by closing the target with the
% computed holes.
%
@(2)@ we take the first @k@ traces returned by @eval@ on the target $\eapp{e}{vs}$
and
@(3)@ we extract the substitutions corresponding to the $\stuck$ traces,
and use them to return the list of witnesses.
%
We obtain the following corollary of Theorem~\ref{thm:soundness}:

\begin{cor}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If |genWitness k $e$ = $\ $ $\triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu}\ldots$|,
  then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that
  $\steps{\eapp{e}{w_1 \ldots w_n}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{cor}

\begin{figure}[t]
\centering
% $$
% \begin{array}{lcll}
% \genWitness{n}{e} &\defeq& \{ \resolve{e'}{\vsu} \spmid \triple{e'}{\vsu}{\tsu} \in \doeval{\genArgs{e}} \} & \\
% \end{array}
% $$

\begin{mcode}
genWitness :: Nat -> $e$ -> [$e$]
genWitness n $e$ = [$\resolve{\eapp{e}{vs}}{\vsu}$ | $\vsu$ <- $\vsu$s]
  where
   vs  = genArgs $e$                  -- (1)
   res = take n (eval $\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}$) $\ $  -- (2)
   $\vsu$s   = [$\vsu$ | $\triple{\stuck}{\vsu}{\tsu}$ <- res] $\ \ $   -- (3)
\end{mcode}
\caption{Generating Witnesses}
\label{fig:algo-gen-witness}
\end{figure}
